---
title: "Building Cohorts with STROBE Diagrams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Cohorts with STROBE Diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8
)
```

## Introduction

The STROBE (Strengthening the Reporting of Observational Studies in Epidemiology) statement provides guidelines for reporting observational studies. A key component is the flow diagram that shows how the study population was selected. The `sRtr` package provides tools to create STROBE-compliant flow diagrams while building your cohort.

This vignette demonstrates how to use the STROBE functions in `sRtr` to:

1. Initialize a cohort with inclusion criteria  
2. Apply sequential filters with exclusion tracking  
3. Generate and customize STROBE flow diagrams  
4. Review the complete filtering log

## Getting Started

```{r setup, eval=FALSE}
library(sRtr)
library(dplyr)
```

## Basic STROBE Workflow

The STROBE workflow in `sRtr` follows these steps:

1. **Initialize** the cohort with `strobe_initialize()`  
2. **Filter** sequentially with `strobe_filter()`  
3. **Plot** the diagram with `plot_strobe_diagram()`  
4. **Review** the log with `get_strobe_log()`

### Example: Kidney Transplant Cohort

Let's build a cohort of kidney transplant recipients with specific criteria:

```{r basic_example, eval=FALSE}
TX_KI %>%
  strobe_initialize(inclusion_label = "All kidney transplants") %>%
  strobe_filter(
    condition = "DON_AGE > 18",
    inclusion_label = "Donor over 18 years old",
    exclusion_reason = "Excluded: Donor age ≤ 18"
  ) %>%
  strobe_filter(
    condition = "DON_ABO == 'B'",
    inclusion_label = "Donor blood type B",
    exclusion_reason = "Excluded: Donor blood type is not B"
  ) %>%
  strobe_filter(
    condition = "REC_AGE >= 18 & REC_AGE <= 65",
    inclusion_label = "Recipient age 18-65",
    exclusion_reason = "Excluded: Recipient age outside 18-65 range"
  )
```

### Reviewing the Filtering Log

```{r review_log, eval=FALSE}
# View the complete filtering log
get_strobe_log()
```

The log shows:
- Each filtering step
- Number of records included/excluded at each step
- Cumulative counts
- Exclusion reasons

## Creating STROBE Diagrams

### Basic Diagram

```{r basic_diagram, eval=FALSE}
plot_strobe_diagram()
```

### Customizing Box Dimensions

```{r custom_dimensions, eval=FALSE}
plot_strobe_diagram(
  incl_width_min = 3,
  incl_height = 1,
  excl_width_min = 2.5,
  excl_height = 1,
  lock_width_min = TRUE,
  lock_height = TRUE
)

plot_strobe_diagram(
  incl_width_min = 3,
  excl_width_min = 2.5,
  incl_fontsize = 16,
  excl_fontsize = 14,
  lock_width_min = TRUE
)
```

### Adjusting Font Sizes

```{r custom_fonts, eval=FALSE}
plot_strobe_diagram(
  incl_fontsize = 16,
  excl_fontsize = 14
)
```

### Publication-Ready Diagrams

```{r publication_ready, eval=FALSE}
plot_strobe_diagram(
  incl_width_min = 4,
  excl_width_min = 3,
  incl_height = 1.2,
  excl_height = 1,
  incl_fontsize = 12,
  excl_fontsize = 10,
  lock_width_min = TRUE,
  lock_height = TRUE
)
```

### Exporting Diagrams

```{r export_diagrams, eval=FALSE}
plot_strobe_diagram(
  export_file = "strobe_diagram.png",
  incl_width_min = 4,
  excl_width_min = 3,
  incl_fontsize = 14,
  excl_fontsize = 12
)

plot_strobe_diagram(
  export_file = "strobe_diagram.svg",
  incl_width_min = 4,
  excl_width_min = 3,
  incl_fontsize = 14,
  excl_fontsize = 12
)
```

## Advanced Examples

### Complex Multi-Step Filtering

```{r complex_example, eval=FALSE}
TX_KI %>%
  strobe_initialize(inclusion_label = "All kidney transplants\n(N=XXX)") %>%
  strobe_filter(
    condition = "DON_AGE >= 18 & DON_AGE <= 70",
    inclusion_label = "Donor age 18-70 years",
    exclusion_reason = "Excluded: Donor age\noutside 18-70 range"
  ) %>%
  strobe_filter(
    condition = "REC_AGE >= 18",
    inclusion_label = "Adult recipients\n(≥18 years)",
    exclusion_reason = "Excluded: Pediatric\nrecipients"
  ) %>%
  strobe_filter(
    condition = "TX_TYPE == 'KIDNEY'",
    inclusion_label = "Kidney-only transplants",
    exclusion_reason = "Excluded: Multi-organ\ntransplants"
  ) %>%
  strobe_filter(
    condition = "!is.na(GRAFT_STAT) & !is.na(GRAFT_FAIL_DATE)",
    inclusion_label = "Complete graft\noutcome data",
    exclusion_reason = "Excluded: Missing graft\noutcome data"
  ) %>%
  strobe_filter(
    condition = "GRAFT_FAIL_DATE >= 365 | is.na(GRAFT_FAIL_DATE)",
    inclusion_label = "≥1 year follow-up\nor event",
    exclusion_reason = "Excluded: <1 year\nfollow-up without event"
  )
```

### Presentation-Style Diagram

```{r presentation_diagram, eval=FALSE}
plot_strobe_diagram(
  incl_width_min = 5,
  excl_width_min = 4,
  incl_height = 1.5,
  excl_height = 1.2,
  incl_fontsize = 18,
  excl_fontsize = 16,
  lock_width_min = TRUE,
  lock_height = TRUE
)
```

## Troubleshooting

```{r sizing_issues, eval=FALSE}
plot_strobe_diagram(
  incl_width_min = 4,
  excl_width_min = 3,
  lock_width_min = TRUE
)

plot_strobe_diagram(
  incl_height = 1.5,
  excl_height = 1.2,
  lock_height = TRUE
)
```

## Summary

The STROBE functions in `sRtr` provide a systematic way to:

- Document your cohort selection process
- Create publication-ready flow diagrams
- Ensure transparent reporting of observational studies
- Maintain a complete audit trail of filtering decisions

Key features:

- Flexible box sizing
- Direct export to PNG/SVG
- Customizable styling
- STROBE-compliant layout

## See Also

- `?strobe_initialize`
- `?strobe_filter`
- `?plot_strobe_diagram`
- `?get_strobe_log`
