---
title: "Building Cohorts with STROBE Diagrams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Cohorts with STROBE Diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8
)
```

## Introduction

The STROBE (Strengthening the Reporting of Observational Studies in Epidemiology) statement provides guidelines for reporting observational studies. A key component is the flow diagram that shows how the study population was selected. We encourage use of the companion `strobe` package, which provides tools to create STROBE-compliant flow diagrams while building your cohort.

This vignette demonstrates how to use the `strobe` functions in `sRtr` to:

1. Initialize a cohort with inclusion criteria  
2. Apply sequential filters with exclusion tracking  
3. Generate and customize STROBE flow diagrams  
4. Review the complete filtering log

For more details, visit the [`strobe` package website](https://vagishhemmige.github.io/strobe/).

### Installation

You can install the development version of the `sRtr` package from GitHub using [`devtools`](https://cran.r-project.org/package=devtools) or [`pak`](https://pak.r-lib.org/):


```{r, eval=FALSE}
# Install devtools if needed
install.packages("devtools")

# Install sRtr (and strobe, a suggested package)
devtools::install_github("VagishHemmige/sRtr", dependencies = TRUE)
```

## Getting Started

```{r setup, eval=FALSE}
library(sRtr)
library(dplyr)
library(strobe)
```

## Basic STROBE Workflow

The STROBE workflow in `sRtr` follows these steps:

1. **Initialize** the cohort with `strobe_initialize()`  
2. **Filter** sequentially with `strobe_filter()`  
3. **Plot** the diagram with `plot_strobe_diagram()`  
4. **Review** the log with `get_strobe_log()`

### Example: Kidney Transplant Cohort

Let's build a cohort of kidney transplant recipients with specific criteria:

```{r basic_example, eval=FALSE}
TX_KI %>%
  strobe::strobe_initialize(inclusion_label = "All kidney transplants") %>%
  strobe::strobe_filter(
    condition = "DON_AGE > 18",
    inclusion_label = "Donor over 18 years old",
    exclusion_reason = "Excluded: Donor age ≤ 18"
  ) %>%
  strobe::strobe_filter(
    condition = "DON_ABO == 'B'",
    inclusion_label = "Donor blood type B",
    exclusion_reason = "Excluded: Donor blood type is not B"
  ) %>%
  strobe::strobe_filter(
    condition = "REC_AGE >= 18 & REC_AGE <= 65",
    inclusion_label = "Recipient age 18-65",
    exclusion_reason = "Excluded: Recipient age outside 18-65 range"
  )
```

### Reviewing the Filtering Log

```{r review_log, eval=FALSE}
# View the complete filtering log
strobe::get_strobe_log()
```

The log shows:
- Each filtering step
- Number of records included/excluded at each step
- Cumulative counts
- Exclusion reasons

## Creating STROBE Diagrams

```{r custom_dimensions, eval=FALSE}
strobe::plot_strobe_diagram(
  incl_width_min = 3,
  incl_height = 1,
  excl_width_min = 2.5,
  excl_height = 1,
  lock_width_min = TRUE,
  lock_height = TRUE,
  incl_fontsize = 16,
  excl_fontsize = 14

)

```
