---
title: "Calculating geographic coverage gain"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating geographic coverage gain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Geography matters greatly in solid organ transplantation.  Donor and recipient practices vary greatly by region, and distance-related barriers can magnify geographic disparities.  In this vignette, we demonstrate how the `calculate_geography_coverage_gain()` function can be used to estimate the number of people living with HIV (PLWH) who gain access to heart transplant centers, defined as residence within a 50-mile radius.


```{r setup, message=FALSE}
library(sRtr)
library(CDCAtlas)
library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(ggplot2)
```

## Create geography objects

We begin by using the `get_hrsa_transplant_centers()` function from the `sRtr` package to create an `sf` object containing heart transplant centers not in PR or HI.  We subsequently perform a substantial amount of data cleaning

```{r}

  Transplant_centers_all_sf<-sRtr::get_hrsa_transplant_centers()%>%
    filter(OPO_STATE_ABBR != "PR" & OPO_STATE_ABBR!= "HI")%>%
    select(OTC_NM, OTC_CD, Service_Lst, X, Y) %>%
    filter(str_detect(Service_Lst, str_to_sentence("Heart")))%>%
    select(-Service_Lst)%>%
    distinct()%>%
    rename(OTCName=OTC_NM, OTCCode=OTC_CD, Latitude=Y, Longitude=X)%>%
    st_transform(5070)  
```

We use the `tigris` package to import a map of states

```{r}

state_geography <- tigris::states(
  year = 2022,
  cb = TRUE,
  class = "sf")%>%  
  filter(!STUSPS %in% c("AK", "HI", "PR", "GU", "VI", "MP", "AS")) %>%
  st_transform(5070)


```


We use ggplot to plot the centers on a map of the United States:

```{r}

ggplot()+
  geom_sf(data=state_geography)+
  geom_sf(data = Transplant_centers_all_sf)+
  theme_minimal()

```

## Define existing centers and additional centers

For the purposes of this analysis, we assume baseline coverage with two centers, *ALUA* (University of Alabama-Birmingham) and *NYMA* (Montefiore Medical Center).

```{r}

existing_centers<-Transplant_centers_all_sf%>%
  filter(OTCCode %in% c("ALUA", "NYMA"))

ggplot()+
  geom_sf(data=state_geography)+
  geom_sf(data = existing_centers, color="red")+
  theme_minimal()


```


We study the additional coverage for PLWH to add *NYCP* (Columbia-Presbyterian) and *TXHI* (Texas Heart Institute at Baylor St. Luke's Medical Center in Houston, TX).


```{r}

new_centers<-Transplant_centers_all_sf%>%
  filter(OTCCode %in% c("NYCP", "TXHI"))

ggplot()+
  geom_sf(data=state_geography)+
  geom_sf(data = new_centers, color="red")+
  theme_minimal()



```


We now create 50-mile buffers (80467.2 meters) around the transplant centers in the previous objects


```{r}
Transplant_centers_all_buffer<-st_buffer(Transplant_centers_all_sf,
                                        dist =80467.2)

existing_centers_buffer<-st_buffer(existing_centers,
                                   dist =80467.2)

ggplot()+
  geom_sf(data=state_geography)+
  geom_sf(data = existing_centers_buffer, color="blue")+
  geom_sf(data = existing_centers, color="red")+
  theme_minimal()


new_centers_buffer<-st_buffer(new_centers,
                              dist =80467.2)

ggplot()+
  geom_sf(data=state_geography)+
  geom_sf(data = new_centers_buffer, color="red")+
  geom_sf(data = new_centers, color="blue")+
  theme_minimal()


  
```

### Importing HIV data

we use the `CDCAtlas` package to import extrapolated estimates of the census tract-level population of PLWH, merging with tract geography obtained via the `tigris` package

```{r}

hiv_tract_data<-CDCAtlas::get_atlas(
  disease = "hiv",
  year = 2022,
  geography = "county",
  extrapolate_to_tract = TRUE
)%>%
  filter(indicator=="HIV prevalence")

tract_geography <- tigris::tracts(
  year = 2022,
  cb = TRUE,
  class = "sf"
) %>%
  select(
    tract_fips = GEOID,
    geometry
  ) %>%
  st_transform(5070)

hiv_tract_sf <- tract_geography %>%
  left_join(
    hiv_tract_data,
    by = "tract_fips"
  )

```


### Estimating additional coverage gain

We finally use the `calculate_geography_coverage_gain` function to estimate the number of PLWH who would gain geographic proximity to a transplant center given the additional centers.

```{r}

calculate_geography_coverage_gain(
    existing_areas=existing_centers_buffer,
    added_areas=new_centers_buffer,
    population_geography=hiv_tract_sf,
    coverage_vars="tract_cases",
    geography_id = NULL,
    method = "intersection",
    crs = 5070,
    output = "wide"
)

```

We see that the additional two centers lead to an improvement from 142,598 PLWH to 177,917, a gain of 35318.  The total number of PLWH in the `hiv_tract_sf` object is 1,099,805. 
