---
title: "Getting Started with sRtr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with sRtr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The `sRtr` package provides comprehensive tools for working with Scientific Registry of Transplant Recipients (SRTR) Standard Analysis Files (SAFs) in R. This vignette will guide you through the essential steps of setting up your environment and loading SRTR data files with proper labeling.

## What is SRTR?

The Scientific Registry of Transplant Recipients (SRTR) is a comprehensive database that tracks transplant data in the United States. The SRTR Standard Analysis Files (SAFs) contain de-identified patient-level data that researchers can use to study transplant outcomes, waiting list dynamics, and donor characteristics.

## Package Overview

The `sRtr` package streamlines the process of:

- Setting up your SRTR data directory
- Loading SRTR files with automatic format detection
- Applying official SRTR variable labels and factor conversions
- Preparing data for analysis with proper documentation

# Installation

You can install the development version of `sRtr` from GitHub:

```{r installation}
# install.packages("devtools")
devtools::install_github("VagishHemmige/sRtr")
```

Load the package:

```{r setup}
library(sRtr)
library(dplyr)
library(labelled)
```

# Setting Up Your SRTR Data Directory

Before you can load SRTR files, you need to tell the package where your SRTR data files are located. The `set_SRTR_wd()` function allows you to specify the directory containing your SRTR files.

## Temporary Setup (Session Only)

For a single R session, you can set the working directory temporarily:

```{r temp_setup}
# Point to your SRTR data directory for this session only
set_SRTR_wd("path/to/your/srtr/files")
```

## Permanent Setup

If you frequently work with SRTR data, you can set the directory permanently by adding the `permanent = TRUE` argument:

```{r permanent_setup}
# Set the directory permanently across R sessions
set_SRTR_wd("path/to/your/srtr/files", permanent = TRUE)
```

This will save your preference and automatically load it in future R sessions.

## Verifying Your Setup

After setting your working directory, the package will automatically detect the available SRTR files in your directory. You can verify that your setup is working correctly by checking if your files are recognized.

# Loading SRTR Files

The primary function for loading SRTR data is `load_srtr_file()`. This function can load both SAS (.sas7bdat) and Parquet (.parquet) files and automatically applies the appropriate SRTR data dictionary.

## Basic File Loading

```{r basic_loading}
# Load a transplant file (liver transplants)
tx_li <- load_srtr_file("TX_LI")

# Load a waiting list file (kidney candidates)
cand_kida <- load_srtr_file("CAND_KIDA")

# Load a donor file
donor <- load_srtr_file("DONOR")
```

## Loading with Labels

One of the key features of `sRtr` is its ability to automatically apply variable labels and factor conversions using the official SRTR data dictionary.

### Variable Labels

Variable labels provide descriptive names for each column:

```{r var_labels}
# Load with variable labels
tx_li <- load_srtr_file("TX_LI", var_labels = TRUE)

# View variable labels
var_label(tx_li$DON_RACE)
var_label(tx_li$REC_AGE_AT_TX)
var_label(tx_li$GRAFT_STAT)
```

### Factor Labels

Factor labels convert coded values to meaningful categories:

```{r factor_labels}
# Load with factor labels
tx_li <- load_srtr_file("TX_LI", factor_labels = TRUE)

# View factor levels
str(tx_li$DON_RACE)
levels(tx_li$DON_RACE)

# View factor labels for transplant status
str(tx_li$GRAFT_STAT)
```

### Complete Labeling

For the most informative data, load with both variable and factor labels:

```{r complete_labels}
# Load with both variable and factor labels
tx_li <- load_srtr_file("TX_LI", var_labels = TRUE, factor_labels = TRUE)

# Now you have both descriptive variable names and meaningful factor levels
str(tx_li$DON_RACE)
var_label(tx_li$DON_RACE)
```

# Working with Labeled Data

Once you've loaded your SRTR data with labels, you can work with it like any other R data frame, but with the added benefit of meaningful variable and factor labels.

## Exploring Your Data

```{r explore_data}
# Basic data exploration
dim(tx_li)
names(tx_li)

# View the first few rows
head(tx_li)

# Summary statistics with meaningful labels
summary(tx_li$DON_RACE)
summary(tx_li$REC_AGE_AT_TX)
```

## Preserving Labels During Analysis

The `labelled` package helps preserve your variable labels during data manipulation:

```{r preserve_labels}
# Filter data while preserving labels
recent_tx <- tx_li %>%
  filter(TX_DATE >= as.Date("2020-01-01")) %>%
  select(TX_DATE, DON_RACE, REC_AGE_AT_TX, GRAFT_STAT)

# Labels are preserved
var_label(recent_tx$DON_RACE)
```

# Independent Labeling Functions

If you already have loaded SRTR data (e.g., from another source), you can apply labels independently using the dedicated labeling functions.

## Applying Factor Labels

```{r apply_factors}
# Load raw data
df <- read_sas("TX_KI.sas7bdat")

# Apply factor labels
df <- apply_srtr_factors(df, filekey = "TX_KI")

# View the result
str(df$DON_RACE)
```

## Applying Variable Labels

```{r apply_varlabels}
# Apply variable labels
df <- apply_srtr_varlabels(df, filekey = "TX_KI")

# View variable labels
var_label(df$DON_RACE)
var_label(df$REC_AGE_AT_TX)
```

## Automatic File Detection

If your data frame name matches the SRTR file name, you can omit the `filekey` parameter:

```{r auto_detect}
# Load data with matching name
TX_KI <- read_sas("TX_KI.sas7bdat")

# Apply labels without specifying filekey
TX_KI <- TX_KI %>%
  apply_srtr_factors() %>%
  apply_srtr_varlabels()
```

# Common SRTR File Types

The `sRtr` package supports all standard SRTR file types. Here are some commonly used files:

## Transplant Files
- `TX_LI` - Liver transplants
- `TX_KI` - Kidney transplants
- `TX_HR` - Heart transplants
- `TX_LU` - Lung transplants
- `TX_IN` - Intestine transplants
- `TX_PA` - Pancreas transplants

## Candidate Files
- `CAND_LIVA` - Liver candidates
- `CAND_KIDA` - Kidney candidates
- `CAND_THORA` - Heart candidates
- `CAND_LUNGA` - Lung candidates

## Other Files
- `DONOR` - Donor information
- `DONOR_DECEASED` - Deceased donor information
- `DONOR_LIVING` - Living donor information

```{r file_examples}
# Examples of loading different file types
liver_tx <- load_srtr_file("TX_LI", var_labels = TRUE, factor_labels = TRUE)
kidney_cand <- load_srtr_file("CAND_KIDA", var_labels = TRUE, factor_labels = TRUE)
donors <- load_srtr_file("DONOR", var_labels = TRUE, factor_labels = TRUE)
```

# Best Practices

## 1. Always Use Labels

Always load your data with both variable and factor labels for the most informative analysis:

```{r best_practice_labels}
# Recommended approach
data <- load_srtr_file("TX_LI", var_labels = TRUE, factor_labels = TRUE)
```

## 2. Check Your Data

Always examine your data after loading to ensure it loaded correctly:

```{r best_practice_check}
# Check dimensions
dim(data)

# Check variable labels
var_label(data[1:5])

# Check factor levels for key variables
str(data$DON_RACE)
str(data$GRAFT_STAT)
```

## 3. Document Your File Sources

Keep track of which SRTR files you're using and their versions:

```{r best_practice_document}
# Good practice: document your data sources
# TX_LI file loaded from SRTR SAF 2024Q1
# Contains liver transplant data from 1987-2023
tx_li <- load_srtr_file("TX_LI", var_labels = TRUE, factor_labels = TRUE)
```

# Troubleshooting

## Common Issues

### File Not Found
If you get a "file not found" error, check:
- Is your working directory set correctly with `set_SRTR_wd()`?
- Are your files in the specified directory?
- Are you using the correct file name (without extension)?

### Labels Not Applied
If labels aren't being applied:
- Ensure you're using the correct file key
- Check that your file name matches the expected SRTR convention
- Verify you're using `var_labels = TRUE` and/or `factor_labels = TRUE`

### Memory Issues
For large files:
- Consider loading only the variables you need
- Work with data in chunks if necessary
- Ensure you have sufficient RAM for your dataset

# Next Steps

Now that you've learned the basics of loading and labeling SRTR data, you might want to explore:

- **STROBE Analysis Workflows**: Learn how to create CONSORT/STROBE diagrams for your research

For more advanced functionality, including STROBE diagram generation, see the additional vignettes in this package.

