---

title: "Mapping transplant geography"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{Mapping transplant geography}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
-------------------------

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(sRtr)
library(sf)
library(dplyr)
```

## Overview

The `sRtr` package includes helper functions and spatial datasets for working with transplant geography in R.

This vignette introduces basic mapping of:

* HRSA transplant center locations
* OPO service areas
* UNOS/OPTN regions

The examples use `sf`, which is the standard R package for working with vector geographic data in R.

## Coordinate reference systems

Spatial objects in `sRtr` are returned as `sf` objects. For many transplant geography tasks, a projected coordinate reference system is preferable to longitude/latitude because it allows more appropriate distance, area, and overlay calculations.

The package commonly uses NAD83 / Conus Albers, EPSG:5070, for U.S. mapping workflows.

```{r crs-example}
sf::st_crs(UNOS_regions_sf)
```

## Built-in spatial datasets

The package includes state-level UNOS/OPTN region geometries and transplant center point locations.

```{r built-in-data}
UNOS_regions_sf
transplant_centers_sf
```

## Plotting UNOS/OPTN regions

The object `UNOS_regions_sf` contains state-level geometries with a `Region` field. These can be plotted directly.

```{r plot-unos-regions}
plot(UNOS_regions_sf["Region"])
```

To plot dissolved region boundaries, use `get_hrsa_optn_regions()`.

```{r plot-dissolved-unos-regions}
optn_regions <- get_hrsa_optn_regions()

plot(optn_regions["Region"])
```

## Plotting transplant centers

The package can download transplant center locations from the HRSA ArcGIS REST service.

```{r get-centers, eval=FALSE}
centers <- get_hrsa_transplant_centers()
```

For a vignette, it is often better to use the built-in example data rather than requiring an internet connection.

```{r plot-centers}
plot(sf::st_geometry(UNOS_regions_sf), border = "grey70")
plot(sf::st_geometry(transplant_centers_sf), add = TRUE, pch = 16, cex = 0.5)
```

## Downloading current HRSA transplant center locations

To retrieve the current HRSA transplant center layer:

```{r current-centers, eval=FALSE}
centers <- get_hrsa_transplant_centers()

centers |>
  dplyr::select(OTC_NM, OTC_CITY, STATE_ABBR, OPO_PROVIDER_NM, Service_Lst)
```

A simple map:

```{r current-centers-map, eval=FALSE}
plot(sf::st_geometry(optn_regions), border = "grey70")
plot(sf::st_geometry(centers), add = TRUE, pch = 16, cex = 0.5)
```

## Plotting OPO service areas

The HRSA service also includes Organ Procurement Organization service area geography.

The detailed OPO service-area layer can be downloaded with:

```{r get-opo-service-areas, eval=FALSE}
opo_areas <- get_hrsa_opo_service_areas()
```

A basic plot:

```{r plot-opo-service-areas, eval=FALSE}
plot(opo_areas["OPO_PROVIDER_NM"])
```

For a less cluttered map, draw only the boundaries and overlay transplant centers.

```{r plot-opo-with-centers, eval=FALSE}
plot(sf::st_geometry(opo_areas), border = "grey70")
plot(sf::st_geometry(centers), add = TRUE, pch = 16, cex = 0.4)
```


## Adding state outlines with tigris

The `tigris` package is useful for adding Census boundary layers, such as states or counties.

```{r tigris-example, eval=FALSE}
library(tigris)

states <- tigris::states(cb = TRUE, year = 2023) |>
  sf::st_transform(sf::st_crs(optn_regions))

plot(sf::st_geometry(states), border = "grey80")
plot(sf::st_geometry(optn_regions), add = TRUE, border = "black", lwd = 2)
plot(sf::st_geometry(transplant_centers_sf), add = TRUE, pch = 16, cex = 0.4)
```

## Filtering by organ or service list

The HRSA transplant center layer includes a `Service_Lst` field. Depending on the analysis, this can be used to identify centers that list specific transplant services.

```{r filter-service-list, eval=FALSE}
kidney_centers <- centers |>
  dplyr::filter(grepl("Kidney", Service_Lst, ignore.case = TRUE))

plot(sf::st_geometry(optn_regions), border = "grey70")
plot(sf::st_geometry(kidney_centers), add = TRUE, pch = 16, cex = 0.5)
```

## Working with geometries

Because these are `sf` objects, they can be used with standard spatial operations.

For example, you can check which transplant centers fall within a UNOS/OPTN region.

```{r spatial-join, eval=FALSE}
centers_with_region <- sf::st_join(
  centers,
  optn_regions |> dplyr::select(Region)
)

centers_with_region |>
  sf::st_drop_geometry() |>
  dplyr::count(Region, sort = TRUE)
```

## Notes

The HRSA transplant center and OPO functions download public GIS layers from HRSA. These layers may change over time as HRSA updates the source data.

The built-in `UNOS_regions_sf` object is included with the package and is useful for reproducible examples and basic region mapping.
